\chapter[%
    Introduction and Literature Review
]{%
    Introduction and Literature Review
}

\section{Naive Electronic Structure Theory}

As an entry point into electronic structure theory, let us begin by forgetting
what we know about electrons from the standard model of particle physics.
From the standpoint of Heisenberg and others developing the new quantum theory
in 1925\cite{Heisenberg:1925p879} chemical matter was described by nimble,
negatively-charged electrons orbiting heavy, positively-charged nuclei.
This theory would be conceptually clarified in the following year by
Schr\"odinger's development of wave mechanics,
\cite{Schrodinger:1926p361,Schrodinger:1926p489,Schrodinger:1926p734} which
described the possible states of electrons in a molecule as eigenfunctions of
the quantum-mechanical Hamiltonian, oscillating in time with a frequency
proportional to their energy.
In atomic units:
\begin{equation}
    \Psi(t)
    =
    \Psi
    e^{-iEt}
    \qquad
    \hat{H}
    \Psi
    =
    E
    \Psi.
\end{equation}
Crudely speaking, this Hamiltonian is derived from its classical counterpart by
replacing its momentum variables with del operators divided by the imaginary
unit,
\(
    \hat{\mathbf{p}}
    =
    \frac{1}{i}
    \nabla
\).
It can be written as a sum over one- and two-electron terms
\begin{equation}
    \hat{H}
    =
    \sum_i^\text{electrons}
    \hat{h}_i
    +
    \sum_{i<j}^{\substack{\text{electron}\\\text{pairs}}}
    \hat{g}_{ij}
\end{equation}
where the one-electron operator
\(
    \hat{h}_i
\)
describes the kinetic energy of the \(i^\text{th}\) electron and its
electrostatic (Coulomb's law) attraction to the nuclei, and the two-electron
operator
\(
    \hat{g}_{ij}
\)
describes the Coulombic repulsion between electrons \(i\) and \(j\).
\begin{equation}
    \hat{h}_i
    \equiv
    \tfrac{1}{2}
    \hat{\mathbf{p}}_i^2
    -
    \sum_A^\text{nuclei}
    \frac{Z_A}{|\mathbf{r}_A - \mathbf{r}_i|}
    \qquad
    \hat{g}_{ij}
    \equiv
    \frac{1}{|\mathbf{r}_i - \mathbf{r}_j|}
\end{equation}
The vector space containing the wavefunction is the system's Hilbert space,
\(\mathcal{H}\), which in our case is the space of square integrable functions
of \(n\) position variables,
\(
    L^2(\mathbb{R}^{3n})
\),
one for each electron in the molecule.
Since this space is infinite-dimensional, the only way forward in most cases is
to determine a basis that approximately spans the states of interest.
Here we are helped by the fact that a function of \(n\) position variables can
be written as a linear combination of products
\begin{equation}
    \Psi(\mathbf{r}_1, \dots, \mathbf{r}_n)
    =
    \sum_{p_1\cdots p_n}
    c_{p_1\cdots p_n}\,
    \phi_{p_1}(\mathbf{r}_1)
    \cdots
    \phi_{p_n}(\mathbf{r}_n)
\end{equation}
where the one-electron functions or {\itshape orbitals} in this expansion span a
one-electron Hilbert space,
\(
    \mathcal{H}_\mathrm{e}
    =
    L^2(\mathbb{R}^3)
\).
More generally, the Hilbert space of a system is always given by a tensor
product of the Hilbert spaces for its constituent particles.
\begin{equation}
    \mathcal{H}
    =
    \mathcal{H}_\mathrm{e}
    \otimes
    \mathcal{H}_\mathrm{e}
    \otimes
    \cdots
\end{equation}
In words, this says that the states of a system are spanned by all possible
combinations of the states that its components exhibit in isolation.
This enables a general strategy for bootstrapping the electronic structure
problem:
\begin{enumerate}
    \item
        \label{item:atomic-orbitals}
        The states of a single electron orbiting a single nucleus are determined
        by the spherical symmetry of the nuclear potential.
        These 1s, 2s, 2p, \dots\ atomic orbitals have analytically known
        functional forms, and their energy increases with the number of nodal
        surfaces in the function.
    \item
        \label{item:mean-field}
        Treating the other electrons in a molecule as static fields which
        partially shield the nuclear charges, the states of individual electrons
        in a molecule can be described as linear combinations of atomic
        one-electron functions centered on each nucleus.
        The dominant contributions to these molecular orbitals come from atomic
        orbitals with a similar energy, so that the lowest-energy molecular
        orbital typically looks like a 1s orbital centered on the nucleus of
        greatest charge, with very little contribution from, say, the 42g
        orbital of another nuclear center.
        The higher energy states are linear combinations of higher energy atomic
        orbitals with increasing numbers of nodes.
    \item
        \label{item:full-ci}
        Finally, the total electronic wavefunction is described by a linear
        combination of products of these molecular orbitals, and the electronic
        ground-state will be spanned to a good approximation by products of the
        low-energy molecular orbitals.
        When the molecular orbitals are weakly interacting, the mean-field
        approximation described in step~\ref{item:mean-field} furnishes a good
        description of their motions, and the wavefunction will be heavily
        dominated by a single one of these products.
        From a statistical perspective, this means that the electron probability
        density of the wavefunction approximately separates into a product of
        one-electron densities, because the orbitals are {\itshape weakly
        correlated} for this state.
\end{enumerate}
At the end of this procedure, we have a matrix equation in the product basis.
\begin{equation}
    \mathbf{H}\mathbf{c}
    =
    E\mathbf{c}
    \qquad
    (\mathbf{H})_{PQ}
    =
    \langle \phi_{p_1}\cdots \phi_{p_n}|
    \hat{H}
    |\phi_{q_1}\cdots \phi_{q_n} \rangle
\end{equation}
In the limit of a complete expansion, this eigenvalue equation is exactly
equivalent to the Schr\"odinger equation and the coefficients of the solution
vector correspond to the components of the wavefunction along our product
functions.
In order to stack the deck in favor of one product in the expansion, we can
determine the molecular orbitals in step~\ref{item:mean-field} to minimize the
energy expectation value
\begin{equation}
    \label{eq:introduction:orbital-product-expectation-value}
    \langle \phi_1\cdots \phi_n|
    \hat{H}
    |\phi_1\cdots \phi_n \rangle
    =
    \sum_{i=1}^\mathrm{orbitals}
    h_i^i
    +
    \sum_{i<j}^{\substack{\mathrm{orbital}\\\mathrm{pairs}}}
    g_{ij}^{ij}
\end{equation}
where we have defined the ubiquitous {\itshape one- and two-electron integrals}.
\begin{equation}
    h_p^q
    \equiv
    \langle\phi_p|\hat{h}_1|\phi_q\rangle
    =
    \int
    d^3\mathbf{r}_1\,
    \phi_p^*(\mathbf{r}_1)
    \hat{h}_1
    \phi_q(\mathbf{r}_1)
\end{equation}
\begin{equation}
    g_{pq}^{rs}
    \equiv
    \langle\phi_p\phi_q|\hat{g}_{12}|\phi_r\phi_s\rangle
    =
    \int
    d^3\mathbf{r}_1
    d^3\mathbf{r}_2\,
    \phi_p^*(\mathbf{r}_1)
    \phi_q^*(\mathbf{r}_2)
    \hat{g}_{12}
    \phi_r(\mathbf{r}_1)
    \phi_s(\mathbf{r}_2)
\end{equation}
For an orthonormal orbital basis, this minimization leads to an effective
Schr\"odinger equation for one-electron states
\begin{equation}
    \label{eq:introduction:mean-field-orbitals}
    (
        \hat{h}_1
        +
        \hat{v}_1
    )
    \phi_i(\mathbf{r}_1)
    =
    \epsilon_i\,
    \phi_i(\mathbf{r}_1)
\end{equation}
where \(\hat{v}_1\) is the mean electrostatic field of the other electrons in
the molecule.
\begin{equation}
    \hat{v}_1
    \equiv
    \sum_{j\neq i}
    \int
    d^3\mathbf{r}_2\,
    \frac{%
        \phi_j^*(\mathbf{r}_2)\phi_j(\mathbf{r}_2)
    }{%
        |\mathbf{r}_1 - \mathbf{r}_2|
    }
\end{equation}
We complete the bootstrapping cycle by expanding these unknown molecular
orbitals in the basis of the atomic orbitals described in
step~\ref{item:atomic-orbitals} and solving for the expansion coefficients.

The general strategy we have just outlined carries over into modern electronic
structure theory, but it is missing two essential ingredients: the spin of the
electron, and the antisymmetric permutational symmetry of electrons as
indistinguishable particles.


\section{Spin}

The non-relativistic theory of one-electron states was completed with Pauli's
solution of the hydrogen spectrum at the start of 1926.\cite{Pauli:1926p336}
His work concludes with a discussion of the recent work by Goudsmit and
Uhlenbeck\cite{Uhlenbeck:1925p953} showing that the anomalous Zeeman splitting
of alkali metals could be explained by positing an intrinsic source of angular
momentum and magnetism for the electron besides that generated by its orbital
motion about the nucleus.
This was the electron's spin.
The need for this quantum number had already been understood by Pauli in his
analysis of alkali metal spectra at the end of 1924:
\begin{quote}
    In alkali metals the angular momentum values of the atom, and its energy
    changes in the presence of an external magnetic field, are appropriately
    interpreted as the sole working of the optically active electron, and the
    same situation is thought to be the case in observations of the anomalous
    Zeeman effect.
    From this standpoint, the doublet structure of the alkali
    spectra, as well as the breakdown of Larmor's theorem, must
    therefore come from some intrinsic, classically non-describable
    type of two-valuedness that is a characteristic of the optically
    active electron.\cite{Pauli:1925p373}
\end{quote}
In hindsight, the Stern-Gerlach experiment\cite{Gerlach:1922p349} had already
shown in 1922 that the 5s electron of the silver atom was quantized into two
magnetic states, whereas the new quantum theory predicted an odd number of
states (\(0, \pm1, \pm2, \dots\)) for the spatial orbits of a charged particle.
This new source of angular momentum was characterized by half-integer values
whose eigenfunctions cannot exist in \(L^2(\mathbb{R}^3)\).
\begin{equation}
    \hat{s}_z
    \psi
    =
    \pm
    \tfrac{1}{2}
    \psi
\end{equation}
The new ``spinor'' part of the electron's state lives in a two-component vector
space
\begin{equation}
    \hat{s}_z
    =
    \begin{pmatrix}
        \tfrac{1}{2} & \hphantom{+}0 \\
        0 & -\tfrac{1}{2}
    \end{pmatrix}
    \qquad
    \alpha
    =
    \begin{pmatrix}
        1 \\ 0
    \end{pmatrix}
    \qquad
    \beta
    =
    \begin{pmatrix}
        0 \\ 1
    \end{pmatrix}
\end{equation}
where \(\alpha\) is the ``up'' spinor and \(\beta\) is the ``down'' spinor.
The states of individual electrons are therefore described not by orbitals but
by {\itshape spin-orbitals}:
\begin{equation}
    \psi(\mathbf{r}, \sigma)
    =
    \phi(\mathbf{r})\,
    \omega_\sigma
    \qquad
    \omega_\sigma
    \equiv
    \left\{
        \begin{array}{cl}
            \alpha_{\sigma}
            &
            \text{if the spin projection is \(+\tfrac{1}{2}\)}
            \\[10pt]
            \beta_{\sigma}
            &
            \text{if the spin projection is \(-\tfrac{1}{2}\)}
        \end{array}
    \right.
\end{equation}
which live in an extended one-electron Hilbert space,
\(
    \mathcal{H}_\mathrm{e}
    =
    L^2(\mathbb{R}^3)
    \otimes
    \mathbb{C}^2
\).
The new spin variable \(\sigma\) refers to the first or second vector component
of the spinor, which is 1 or 0 depending on whether the state is spin-up or
spin-down.

Having completed the system of quantum numbers for an electron in a spherical
potential with what would eventually be recognized as spin, Pauli was struck
with a curious observation:
\begin{quote}
    By considering the case of strong magnetic fields we can reduce [earlier
    observations], that the number of electrons in a completed subgroup is the
    same as the number of corresponding terms in the Zeeman effect of the alkali
    spectra, to the following more general rule about the occurrence of
    equivalent electrons in an atom:
    There can never be two or more equivalent electrons in an atom for which
    in strong fields the values of all quantum numbers \(n, l, k, m_l\) (or,
    equivalently, \(n, l, m_l, m_s\)) are the same.
    If an electron is present in the atom for which these quantum numbers
    (in an external field) have definite values, this state is ``occupied.''
    \dots\
    We cannot give a further justification for this rule, but it seems to be a
    very plausible one.\cite{Pauli:1925p756}
\end{quote}
This ``housing office for equivalent orbits''\cite{Mehra:1982} would remain a
mystery until Heisenberg's work on two-electron systems the following year.


\section{Antisymmetry}

In June of 1926, Heisenberg published an article on {\itshape The Many-Body
Problem and Resonance in Quantum Mechanics} in which he aimed to ``lay the
foundations for a quantum mechanical treatment of many-body
systems.''\cite{Heisenberg:1926p411}
He sought to reconcile three outstanding problems in the quantum theory of
electrons.
In his words:
\begin{quote}
    [1.] The aspects of de Broglie's theory of waves that lead to Bose-Einstein
    statistics appear to have no analogue in quantum mechanics;
    [2.] Ad hoc rules like Pauli's ban on equivalent orbitals cannot be
    expressed in the current mathematical formalism of quantum mechanics\dots\ 
    [3.] Finally there is one known difficulty in the quantitative
    interpretation of spectra that we should remind ourselves of:
    The splitting of singlet and triplet states in the spectra of the alkaline
    earth metals and in helium is too big by an order of magnitude to be
    explained as a difference in the magnetic interaction energies of two
    spinning electrons.\cite{Heisenberg:1926p411}
\end{quote}
By considering two quantum harmonic oscillators and coupling them as
indistinguishable particles, in line with the statistical considerations of Bose
and Einstein,\cite{Bose:1924p178,Einstein:1924p261,Einstein:1925p3} Heisenberg
found that the eigenstates exist in symmetric or antisymmetric combinations, and
that only the antisymmetric combinations are consistent with the observed
spectrum of helium.
That is, the electronic wavefunction appears to be antisymmetric under particle
exchange, so that the appropriate basis states are not simple products but
antisymmetric ones\footnote{%
    The arguments here are short-hand for the space and spin coordinates of the
    electrons, as in \((1) \equiv (\mathbf{r}_1, \sigma_1)\).
}
\begin{equation}
    \Phi_{p_1\cdots p_n}(1, \ldots, n)
    =
    \tfrac{1}{\sqrt{n!}}
    \sum_{\pi}
    (-)^\pi
    \psi_{p_1}(\pi_1)
    \cdots
    \psi_{p_n}(\pi_n)
\end{equation}
where \(\pi=(\pi_1\cdots \pi_n)\) runs over permutations of the electron labels
\((1\cdots n)\) with signature \((-)^\pi\).
Whenever the same orbital appears twice in such an expansion the function
vanishes, so this provided a mathematical basis for Pauli exclusion.
The energy of the antisymmetric product state has the form
\begin{equation}
    \label{eq:introduction:determinant-expectation-value}
    \langle \Phi_{1\cdots n}|
    \hat{H}
    |\Phi_{1\cdots n} \rangle
    =
    \sum_{i=1}^\mathrm{orbitals}
    h_i^i
    +
    \sum_{i<j}^{\substack{\mathrm{orbital}\\\mathrm{pairs}}}
    \overline{g}_{ij}^{ij}
    \qquad
    \overline{g}_{ij}^{ij}
    \equiv
    g_{ij}^{ij}
    -
    g_{ij}^{ji}
\end{equation}
where only the exchange integrals
\(
    g_{ij}^{ji}
    =
    \langle \psi_i\psi_j|\hat{g}_{12}|\psi_j\psi_i\rangle
\)
are new from the single-product expectation value
(\cref{eq:introduction:orbital-product-expectation-value}).
These ``exchange interactions'' between electrons in orbitals \(i\) and \(j\)
serve to lower the energy by cancelling out part of the Coulomb repulsion where
the orbitals overlap.
Moreover, the orthogonality of opposite spinors means that this effect only
takes place between electrons of the same spin, and thereby lowers the energy of
high-spin states relative to low-spin ones of the same configuration, as had
been observed by Hund.\cite{Hund:1925p345}
Thus the new theory could also explain the ``large force of unknown
origin''\cite{Mehra:1982} that was lowering the energy of the triplet states in
helium.


\section{Modern Electronic Structure Theory}

\begin{enumerate}
    \item
        The general framework set forth by Heisenberg remains the basis for
        non-relativistic electronic structure theory, in that our goal is to
        solve the Schr\"odinger equation in the basis of antisymmetric
        spin-orbital products.
        \begin{equation}
            \Psi(1, \ldots, n)
            =
            \sum_{p_1<\cdots<p_n}
            c_{p_1\cdots p_n}
            \Phi_{p_1\cdots p_n}(1, \ldots, n)
        \end{equation}
    \item
        However, a considerable mathematical machinery has been developed to
        allow for a more natural way of dealing with the antisymmetry of these
        functions.
    \item
        Notice that antisymmetry implies that the Hilbert space for a
        many-electron system is actually smaller than we thought.
        Namely, the wavefunction exists in a subspace
        \begin{equation}
            \mathcal{H}_{n\mathrm{e}}
            =
            P(
                \underset{\text{$n$ times}}{%
                    \mathcal{H}_\mathrm{e}
                    \otimes\cdots
                    \otimes
                    \mathcal{H}_\mathrm{e}
                }
            )
        \end{equation}
        where \(P\) denotes the projection onto the span of the antisymmetric
        products.
    \item
        We can define a map from the \(n\)-electron Hilbert space to the
        \(n-1\)-electron Hilbert space as an integral operator
        \begin{equation}
            \hat{a}_p
            \Psi(1, \ldots, n)
            \equiv
            \sqrt{n}
            \int
            d(1)\,
            \psi_p^*(1)\,
            \Psi(1, \ldots, n)
        \end{equation}
        which enables a resolution of the identity along the first coordinate.
        \begin{equation}
            \Psi(1, \ldots, n)
            =
            \tfrac{1}{\sqrt{n}}
            \sum_p
            \psi_p(1)\,
            \hat{a}_p
            \Psi(1, \ldots, n)
        \end{equation}
    \item
        Substituting this expansion into the energy expectation value of the
        wavefunction
        \begin{equation}
            \langle\Psi|\hat{H}|\Psi\rangle
            =
            \sum_{pq}
            h_p^q\,
            \langle a_p\Psi| a_q\Psi\rangle
            +
            \tfrac{1}{4}
            \sum_{pqrs}
            \overline{g}_{pq}^{rs}\,
            \langle a_q a_p\Psi| a_s a_r\Psi\rangle
        \end{equation}
    \item
        Restricted to \(\mathcal{H}_{n\mathrm{e}}\)
        \begin{equation}
            \hat{H}
            =
            h_p^q\,
            a^p_q
            +
            \tfrac{1}{4}
            \overline{g}_{pq}^{rs}
            a^{pq}_{rs}
            \qquad
            a^p_q
            \equiv
            a_p^\dagger a_q
            \qquad
            a^{pq}_{rs}
            \equiv
            a_p^\dagger a_q^\dagger a_s a_r
        \end{equation}
    \item
\end{enumerate}


\section{Cumulants of the Electron Density}

\begin{equation}
    \langle\Psi|Q|\Psi\rangle
    =
    \sum_{k=1}^n
    \sum_{\Pi}^{\mathrm{Part}(Q, k)}
    (-)^{\mathrm{sgn}(\Pi)}
    \langle\Psi|\Pi_1|\Psi\rangle_\mathrm{c}
    \cdots
    \langle\Psi|\Pi_k|\Psi\rangle_\mathrm{c}
\end{equation}

\begin{equation}
    \langle\Psi|Q|\Psi\rangle_\mathrm{c}
    =
    \sum_{k=1}^n
    (-)^{k+1}
    (k-1)!
    \sum_{\Pi}^{\mathrm{Part}(Q, k)}
    (-)^{\mathrm{sgn}(\Pi)}
    \langle\Psi|\Pi_1|\Psi\rangle
    \cdots
    \langle\Psi|\Pi_k|\Psi\rangle
\end{equation}

\begin{equation}
    \gamma^{p_1\cdots p_n}_{q_1\cdots q_n}
    \equiv
    \langle\Psi|a^{p_1\cdots p_n}_{q_1\cdots q_n}|\Psi\rangle
    \qquad
    \lambda^{p_1\cdots p_n}_{q_1\cdots q_n}
    \equiv
    \langle\Psi|a^{p_1\cdots p_n}_{q_1\cdots q_n}|\Psi\rangle_\mathrm{c}
\end{equation}

\begin{equation}
    \gamma^{pqr}_{stu}
    =
    \lambda^{pqr}_{stu}
    +
    P^{(pq/r)}_{(st/u)}
    \lambda^{pq}_{st}
    \gamma^r_u
    +
    P_{(s/t/u)}
    \gamma^p_s
    \gamma^q_t
    \gamma^r_u
\end{equation}

\begin{equation}
    \begin{array}{r@{\,}l}
        \gamma^{pqrs}_{tuvw}
        =&
        \lambda^{pqrs}_{tuvw}
        +
        P_{(tuv/w)}^{(pqr/s)}
        \lambda^{pqr}_{tuv}
        \gamma^s_w
        +
        P_{(tu/vw)}
        \lambda^{pq}_{tu}
        \lambda^{rs}_{vw}
        \\[5pt]
        &
        +
        P_{(tu/v/w)}^{(pq/rs)}
        \lambda^{pq}_{tu}
        \gamma^r_v
        \gamma^s_w
        +
        P_{(t/u/v/w)}
        \gamma^p_t
        \gamma^q_u
        \gamma^r_v
        \gamma^s_w
    \end{array}
\end{equation}


\section{Prospectus}

